---
title: "Chapter 4: Sensitivity Analysis"
header-includes:
- \usepackage{longtable}
- \usepackage{caption}
monofont: Times New Roman
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
    toc: TRUE
    number_sections: true
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
---

\pagestyle{headings}

```{r setup, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy = F)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(width = 30)
library (magrittr, quietly = TRUE) #Used for pipes/dplyr
library (dplyr, quietly = TRUE)
library (ROI, quietly = TRUE)
library (ROI.plugin.glpk, quietly = TRUE)
library (ompr, quietly = TRUE)
library (ompr.roi, quietly = TRUE)
library (pander, quietly = TRUE)
library (gridExtra)
library (formatR)
```

# Linear Programming Sensitivity Analysis

We can get a lot more than just the objective function value and the decision variable from a solved linear program.  In particular, we can potentially explore the impact of changes in constrained resources, changes in the objective function, forced changes in decision variables, and the introduction of additional decision variables.  

## Base Case

To demonstrate this, let's revisit our explicit model of production planning.  We will use the explicit version for the sake of clarity and simplicity but the same techniques could be used for the generalized model or other linear programs.  
\vspace{12pt}

Recall the explicit production planning problem from earlier on Formula 4.1.  

```{marginfigure}
$$
 \begin{split}
 \begin{aligned}
    \text{Max: } & 20Chairs+14Desks+16Tables \\
    \text{s.t.} & \\
         & 6Chairs+2Desks+4Tables \leq 2000 \\
         & 8Chairs+6Desks+4Tables \leq 2000 \\
         & 6Chairs+4Desks+8Tables \leq 1440 \\
         & 40Chairs+25Desks+16Tables \leq 9600 \\
         & Tables \leq 200 \\
         & Chairs,  \; Desks, \; Tables \geq 0  
  \end{aligned}
  \end{split}
$$
```
```{marginfigure}
**Formula 4.1**: Explicit model of production planning
```

The implementation that we did earlier for production planning was straightforward.

```{r Base3VarModel, fig.margin=TRUE,fig.width=7}
Base3VarModel <- MIPModel() %>%
  add_variable(Chairs, type ="continuous", lb = 0) %>%
  add_variable(Desks, type ="continuous",lb = 0) %>%
  add_variable(Tables, type ="continuous", lb = 0) %>%
  
  set_objective(20*Chairs+14*Desks+16*Tables,"max")%>%
  
  add_constraint(6*Chairs+2*Desks+4*Tables<= 2000)%>%
  #fabrication
  add_constraint(8*Chairs+6*Desks+4*Tables<= 2000)%>% 
  #assembly
  add_constraint(6*Chairs+4*Desks+8*Tables<= 1440)%>% 
  #machining
  add_constraint(40*Chairs+25*Desks+16*Tables<= 9600)%>% 
  #wood
  add_constraint(Tables <= 200) %>% #
  solve_model(with_ROI(solver = "glpk"))
  obj_val <- objective_value(Base3VarModel)
  xchairs <- get_solution (Base3VarModel, Chairs)
  xdesks  <- get_solution (Base3VarModel, Desks)
  xtables <- get_solution (Base3VarModel, Tables)
base_case_res <- cbind(objective_value(Base3VarModel),
                    get_solution(Base3VarModel, Chairs),
                    get_solution(Base3VarModel, Desks),
                    get_solution(Base3VarModel, Tables))
colnames(base_case_res)<-list("Profit", 
                              "Chairs", 
                              "Desks", 
                              "Tables")
rownames(base_case_res)<-list("Base Case")
grid.table(base_case_res)
```
```{marginfigure}
**Table 4.1**: Base Case Production Plan
```

In the base case, we are producing chairs and desks but not tables to generate a total profit of $\$4880$.  

## Shadow Prices

There are many resources, some are fully used while some are not fully utilized.  How could we prioritize the importance of each resource?  For example, if the factory manager could add a worker to one department, which should it be? Conversely, if an outside task came up where should she draw the time from?  We could modify the model and rerun.  For complex situations, this may be necessary.  On the other hand, we could also use sensitivity analysis to explore the relative value of the resources.  
\vspace{12pt}

Let's begin by examining the row duals, also known as shadow prices.
```{r rduals_3var,fig.margin=TRUE,fig.width=5}
#rduals1 <-as.matrix(get_row_duals(result1), ncol=5)
rduals1 <-as.matrix(get_row_duals(Base3VarModel))
dimnames(rduals1)<-list(c("fabrication", 
                          "assembly", 
                          "machining", 
                          "wood", 
                          "demand"), 
                          c("Row Duals"))
grid.table(format(rduals1,digits=4))
```
```{marginfigure}
**Table 4.2**: Shadow Prices of Constrained Resources
```
The row duals or shadow prices for fabrication is zero.  This means that the marginal value of one additional hour of fabrication labor time is $\$0$. This makes sense if you examine the amount of fabrication time used and realize that the company is not using all of the 2000 fabrication department's time available.  Therefore adding for fabrication hours certainly can't help improve the production plan.    
\vspace{12pt}

On the other hand, all of the assembly time (resource) is used in the optimal solution.  The shadow price of an hour of assembly time is $\$1.0$.  This means that for every hour of additional assembly time within certain limits, the objective function will increase by $\$1.0$ also.
\vspace{12pt}

All of the 1440 hours of labor available in the machining center are consumed by the optimal production plan.  Increasing the hours available may allow the company to change the production plan and increase the profit.  While you could rerun the model with increased machining hours to determine the new  optimal production plan but if only want to know the change in the optimal objective function value, you can determine that from the shadow price of the machining constraint. Each additional hour (within a certain range) will increase the profit by $\$2.0$.  
\vspace{12pt}

This potential for increased profit can't be achieved by simply increasing the resource - it requires a modified production to utilize this increased resource.  To find the production plan that creates this increased profit, let's solve a modified linear program.

## Testing Adding Labor to Machining

Let's test the numerical results from the Shadow Price table by adding an hour of labor to the machining center. The model is represented in the Formula 4.2.  

```{marginfigure}
$$
 \begin{split}
 \begin{aligned}
    \text{Max:  } & 20Chairs+14Desks+16Tables \\
    \text{s.t.} \\
          & 6Chairs+2Desks+4Tables \leq 2000 \\
          & 8Chairs+6Desks+4Tables \leq 2000 \\
          & 6Chairs+4Desks+8Tables \leq 1441 \\
          & 40Chairs+25Desks+16Tables \leq 9600 \\
          & Tables \leq 200 \\
          & Chairs,  \; Desks, \; Tables \geq 0  
  \end{aligned}
  \end{split}
$$
```
```{marginfigure}
**Formula 4.2**: Explicit model of production planning adding labor to machining
```
\vspace{12pt}
And the code is showed in the following chunk, with Table 4.3 showing the optimal model result.
```{r IncMachiningHrs, fig.margin=TRUE}
IncMachiningHrs <- MIPModel() %>%
  add_variable(Chairs, type = "continuous", lb = 0) %>%
  add_variable(Desks, type = "continuous",lb = 0) %>%
  add_variable(Tables, type = "continuous", lb = 0) %>%
  
  set_objective(20*Chairs + 14*Desks + 16*Tables,"max")%>%
  
  add_constraint(6*Chairs + 2*Desks + 4*Tables<= 2000)%>% 
  #fabrication
  add_constraint(8*Chairs + 6*Desks + 4*Tables<= 2000)%>% 
  #assembly
  add_constraint(6*Chairs + 4*Desks + 8*Tables<= 1441)%>% 
  #machining
  add_constraint(40*Chairs + 25*Desks + 16*Tables<= 9600)%>% 
  #wood
  add_constraint(Tables <= 200) %>% #
  solve_model(with_ROI(solver = "glpk"))
inc_mc_res  <- cbind(objective_value(IncMachiningHrs),
                     get_solution(IncMachiningHrs, Chairs),
                     get_solution(IncMachiningHrs, Desks),
                     get_solution(IncMachiningHrs, Tables))
colnames(inc_mc_res)<-list("Profit", 
                           "Chairs", 
                           "Desks", 
                           "Tables")
rownames(inc_mc_res)<-list("Increased M/C Hrs")
temp1 <- rbind(base_case_res, inc_mc_res)
grid.table(temp1)
```
```{marginfigure}
**Table 4.3**: Production Plan with One Additional Machining Hour
```
\vspace{12pt}
These results confirmed that adding one hour of machining time results in a _new_ production plan that generates a profit of two more dollars, exactly as expected.  
\vspace{12pt}

The Shadow Prices can be very helpful in larger problems where questions might come up of trying to evaluate or prioritize limited resources.  

## Shadow Prices of Underutilzed Resources

The shadow price on wood is zero (as was also the case for assembly hours).  This means that even a large increase in wood would not affect the maximum profit or the optimal production plan. Essentially there is plenty of wood, having more would not allow enable a better profit plan to be possible.  Let's confirm this as well with a numerical example by adding _10,000_ more square feet of wood. See formula 4.3. 
```{marginfigure}
$$
 \begin{split}
 \begin{aligned}
    \text{Max:  }   & 20Chairs+14Desks+16Tables \\
    \text{s.t.} & \\
        & 6Chairs+2Desks+4Tables \leq 2000 \\
        & 8Chairs+6Desks+4Tables \leq 2000 \\
        & 6Chairs+4Desks+8Tables \leq 1440 \\
        & 40Chairs+25Desks+16Tables \leq 19600 \\
        & Tables \leq 200 \\
        & Chairs,  \; Desks, \; Tables \geq 0  
  \end{aligned}
  \end{split}
$$
```  
```{marginfigure}
**Formula 4.3**: Explicit model for Underutilzed Resources
```
\vspace{12pt}
In the same way that it was done in previous point, the next chunk shows how to build in R an underutilized scenario. Table 4.3 contains the result for this model.

```{r IncWood, fig.margin=TRUE}
IncWood<- MIPModel() %>%
  add_variable(Chairs, type = "continuous", lb = 0) %>%
  add_variable(Desks, type = "continuous",lb = 0) %>%
  add_variable(Tables, type = "continuous", lb = 0) %>%
  
  set_objective(20*Chairs + 14*Desks + 16*Tables,"max")%>%
  
  add_constraint(6*Chairs + 2*Desks + 4*Tables<= 2000)%>% 
  #fabrication
  add_constraint(8*Chairs + 6*Desks + 4*Tables<= 2000)%>% 
  #assembly
  add_constraint(6*Chairs + 4*Desks + 8*Tables<= 1440)%>% 
  #machining
  add_constraint(40*Chairs + 25*Desks + 16*Tables<= 19600)%>% 
  #wood
  add_constraint(Tables <= 200) %>% #
  solve_model(with_ROI(solver = "glpk"))
inc_wd_res  <- cbind(objective_value(IncWood),
                     get_solution (IncWood, Chairs),
                     get_solution (IncWood, Desks),
                     get_solution (IncWood, Tables))
colnames(inc_wd_res)<-list("Profit", 
                           "Chairs", 
                           "Desks", 
                           "Tables")
rownames(inc_wd_res)<-list("Increased Wood")
temp2 <- rbind(base_case_res, inc_mc_res, inc_wd_res)
       grid.table(temp2)
```
```{marginfigure}
**Table 4.4**: Production Plan with 10,000 More Wood
```
The demand constraint of 200 is an upper limit only on Tables.  Since far fewer than 200 Tables are produced in the optimal production plan, relaxing this constraint also has no impact (shadow price=0).

## Reduced Costs of Variables

Next, we move on to the *reduced costs* of variables. This concept often requires rereading  several times. The mathematical details rely on the structure of linear programming and the simplex algorithm. We won't go into great detail on the origin of this mathematically in detail here.  
\vspace{12pt}

*Reduced Cost of Chairs*

Let's start by examining the Chairs. The reduced cost for Chairs is the per unit marginal profit minus the per unit value (in terms of shadow prices) of the resources used by a unit in production.  
\vspace{12pt}

Next, let's discuss the column duals which are often referred to as reduced costs of variables.  Let's extract these from the results just as we did for the shadow prices.

```{r cduals_3var, fig.margin=TRUE}
#cduals1 <-as.matrix(get_column_duals(result1), ncol=1 )
cduals1 <-as.matrix(get_column_duals(Base3VarModel) )
dimnames(cduals1)<-list(c("Chairs", "Desks", "Tables"), 
                        c("Column Duals"))
grid.table(format(cduals1,digits=4))
```
```{marginfigure}
**Table 4.5**: Reduced Costs of Variables
```

These results are interesting.  The reduced costs of variables that are between simple upper and lower bounds will be zero. The reduced cost of a variable can be thought as the difference between the value of the  resources consumed by the resources and the value of the variable in the objective function.  All of our variables have simple lower bounds of zero and no upper bounds.  The first two (Chairs and Desks) have a zero reduced cost and the last one (Tables) is negative.  

Let's start by examining the Shadow Prices of the resources along with the number of each resource used in the production of a single chair.  

```{r Shadow_Price_Single_Chair, fig.margin=TRUE}
chair_res_used<-cbind(rduals1,c(6,8,6,40,0)) #The numbers dont match with the equation
colnames(chair_res_used)<-c("Row Duals", "Resources Used")
grid.table(format(chair_res_used, digits=4))
```
```{marginfigure}
**Table 4.6**: Resources Used by a Chair and their Shadow Prices
```
Taking the product in each row and adding them up will give the marginal value of the resources consumed by an incremental change in production of chairs. In this case, the marginal value is $0+1*8+2*6+0+0=20$.  Since the profit per chair (from the objective function coefficient on Chairs) is also $\$ 20$, they are in balance and the difference between them is zero which is why the reduced cost for Chairs is zero. This can be thought of saying that there is marginal benefit of cost to a very small forced change in the value of the Chairs variable.  The value of the resources used in producing a chair equals that of the profit of a chair.  
\vspace{12pt}

We can repeat the same process and calculations for Desks.  

```{r Shadow_Price_Desks,fig.margin=TRUE }
desk_res_used<-cbind(rduals1,c(2,6,4,25,0)) #The numbers dont match with the equation
colnames(desk_res_used)<-c("Row Duals", "Resources Used")
grid.table(format(desk_res_used, digits=4))
```
```{marginfigure}
**Table 4.7**: Resources Used by a Desk and their Shadow Prices
```
In this case, using the columns from Table 4.6 we can calulate the marginal value as $0+1*6+2*4+0+0=14$.  Since the profit per desk (from the objective function coefficient on Desks) is also $\$ 14$, they are in balance and the difference between them is zero which is why the reduced cost for the _Desks_ variable is also zero.

*Reduced Price of Tables*

The situation is more interesting for Tables. The production plan does not call for producing any tables. This means that it is sitting right at the lower bound of zero Tables.  This hints that perhaps we would like to make even less than zero Tables if we could and that being _forced_ to make even one table would be costly. A reduced cost of -4 means that the opportunity cost of resources used in producing a table is $\$4$ more than the profit of producing a single table.  In other words, we would expect that if we force the production of a single table, the over all production plan's profit will go down by $\$4$.
\vspace{12pt}

Let's go ahead and test this interpretation by again looking at the value of the resources consumed in the production of a table and the number used. 

```{r Shadow_Price_Table, fig.margin=TRUE}
table_res_used<-cbind(rduals1,c(4,4,8,16,0))  #The numbers dont match with the equation
colnames(table_res_used)<-c("Row Duals", "Resources Used")
grid.table(table_res_used)
```
```{marginfigure}
**Table 4.8**: Resources Used by a Table and their Shadow Prices
```
Notice that the values based on shadow prices of the resources used by a table are $\$1*4+\$2*8=\$20$. Alas, the profit for each table is just $\$16$ which means that forcing the production of a single table will decrease the production plan's profit by $\$4$. In other words, the impact on the objective function is   $\$-4$ which is the same as the reduced price entry of Tables.  
\vspace{12pt}

Okay, now let's put both side by side in a table to show the results.
\vspace{12pt}

Okay, now let's test it.  We will modify the formulation to set a lower bound on the number of Tables to be 1.  Note that we do this in case by setting the _lb_ option in the *add_variable* to be 1.  Also, the demand constraint for tables could also be accommodated by setting the table variable's upper bound (*ub*) to 200.  

```{r Table1Model, fig.margin=TRUE,fig.width=5}
Table1Model <- MIPModel() %>%
  add_variable(Chairs, type = "continuous", lb = 0) %>%
  add_variable(Desks, type = "continuous",lb = 0) %>%
  add_variable(Tables, type = "continuous", lb = 1) %>%
  
  set_objective(20*Chairs + 14*Desks + 16*Tables,"max")%>%
  
  add_constraint(6*Chairs + 2*Desks + 4*Tables<= 2000)%>% 
  #fabrication
  add_constraint(8*Chairs + 6*Desks + 4*Tables<= 2000)%>% 
  #assembly
  add_constraint(6*Chairs + 4*Desks + 8*Tables<= 1440)%>% 
  #machining
  add_constraint(40*Chairs + 25*Desks + 16*Tables<= 9600)%>% 
  #wood
  add_constraint(Tables <= 200) %>% #
  solve_model(with_ROI(solver = "glpk"))
  obj_val <- objective_value(Table1Model)
  xchairs <- get_solution (Table1Model, Chairs)
  xdesks  <- get_solution (Table1Model, Desks)
  xtables <- get_solution (Table1Model, Tables)
  Table1_case_res           <- cbind(xchairs,
                                     xdesks,
                                     xtables,
                                     obj_val)
  rownames(Table1_case_res) <- ""
  #rownames(Table1_case_res) <- "Amount to Produce"
  grid.table(Table1_case_res)
```
```{marginfigure}
**Table 4.9**: Production Plan with Table Set to One
```
\vspace{12pt}
Let's compare the results for the new production plan and the original base case looking at the Table 4.9.  

```{r Base_vs_Forced_Table, fig.margin=TRUE}
rownames(base_case_res) <- "Base Case"
Table1_case_res <- cbind(obj_val, xchairs,xdesks,xtables)
rownames(Table1_case_res) <- "After change"
temp3 <-rbind(base_case_res,Table1_case_res)
grid.table(temp3)
```
```{marginfigure}
**Table 4.10**: Impact of a Forced Change in Tables
```
As we expected, the forced change of making one additional table resulted in a decrease of the overall profit from  $\$4800$ to  $\$4876$. This occurred because making one table meant that we had fewer of the precious, limited resources. \vspace{12pt}

This meant that the number of chairs and desks were changed resulting lower profit even though a table is on its own profitable.

* Evaluating a New Product Design with Shadow Prices *

Let's consider a design proposal for a bookcase. The bookcase has a projected profit of $\$35$ each and uses 6 hours of fabrication time, 12 of assembly, and 16 of machining. It uses 80 square feet of wood.

```{r New_Prod_with_Shadow_Prices,fig.margin=TRUE}
bookcase_res_used<-cbind(rduals1,c(6,12,16,80,0))
colnames(bookcase_res_used)<-c("Row Duals", 
                               "Resources Used")
grid.table(format(bookcase_res_used,digits=4))
```
```{marginfigure}
**Table 4.10**: Resources Used by a Bookcase and their Shadow Prices
```
Even without adding it to the model, we can check to see if it is worthwhile to consider seriously.  The opportunity cost of producing one bookcase would result in the $\$35 - \$1*12  - \$2*16=\$-9 $.  In other words, even though a bookcase has higher profit than any other product, producing one would cost the company nine dollars of overall profit. 
\vspace{12pt}

We could interpret this as a simple hard stop on the decision to produce bookcases but we could go one step further by setting a target for redesigning the product or its production.  If the machining time could be reduced by 4.5 hours to 11.5 hours, then the value of the resources consumed would be equal to the profit and we would be indifferent to producing some bookcases. 
\vspace{12pt}


## Exercises 


```{exercise, name="Adding Frames"}
```
Your company has extended production to allow for producing picture frames and is now including a finishing department that primes and paints (or stains) the furniture.

| Characteristic | Chairs | Desks | Frames | Tables | Available  | 
|---------------:|:-----:|:-----:|:------:|:---------:|:----------:|
|  Profit        |  $20  |  $14  |    $3  |  $16      |            |
|  Fabrication   |   6   |   2   |     1  |    4      |   1440     |
|  Assembly      |   8   |   6   |     1  |    8      |   1440     |
|  Machining     |   6   |   4   |     1  |   25      |   2000     |
|  Painting      |   7   |  10   |     2  |   12      |   1000     |
|  Wood          |  40   |  25   |     5  |   16      |   9600     |

a) Use R Markdown to create your own description of the model.
b) Extend the R Markdown to show your LP Model.  Be sure to define models.
c) Solve the model in R.
d) Interpret and discuss the model in R Markdown.
e) Examine and reflect upon the reduced costs and shadow prices from the context of which products to produce and not produce.
f) Using the results from e), (i.e. reduced cost and shadow prices) make one change to the base model's **objective function** that will change the production plan.  Rerun and discuss the new results.
g) Using the results from e), (i.e. reduced cost and shadow prices) make one change to the base model's **resource usage values** that will change the production plan. Rerun and discuss the new results.
h) Using the results from e), (i.e. reduced cost and shadow prices) make one change to the base model's **available resource values** that will change the production plan. Rerun and discuss the new results. 
i) Combine the results of the base case e), as well as the variations f) through h) into a single table and discuss the results. 
\vspace{12pt}

```{exercise, name="Revisiting Transportation"}
``` 

Using sensitivity analysis, revisit the transportation exercise from chapter 3.  
a) If one more unit of supply was available, where would it be prioritized and why?  
b) If demand could be increased by one unit, would it affect the result and at which destination node it be preferred and why?
