---
title: "Appendix Introduction to Summation Notation"
author: "Caroline Blackledge, Alexander Keller, and Shahram Khorasanizadeh"
output: 
  tufte::tufte_handout:
    tufte_variant: "envisioned"
  tufte::tufte_html:
    tufte_variant: "envisioned"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy = F)
```

# Appendix: An Introduction to Summation Notation {-}

## Basic Summation Notation

In linear programming, we often need to sum sets of numbers, such as:
1 + 2 + 3 + 4 + 5 + 6 + 7 + 8 + 9 + 10
+ 11 + 12 + 13 + 14 + 15 + 16 + 17 + 18 + 19 + 20
+ 21 + 22 + 23 + 2 + 25 + 26 + 27 + 28 + 29 + 30
+ 31 + 32 + 33 + 34 + 35 + 36 + 37 + 38 + 39 + 40
+ 41 + 42 + 43 + 44 + 45 + 46 + 47 + 48 + 49 + 50

Summation notation is a mathematical shorthand used to represent repeated addition of algebraic expressions involving one or more variables. The Greek capital letter sigma, $\sum$, is the symbol used to show that we wish to make calculations using summation notation. Since summation notation includes at least one variable--let's limit this initial examination to one variable only--the variable is displayed below the $\sum$ symbol. The summation notation below tells us that we will be finding the summation of a variable expression involving $x$.  

$$\sum_{x}$$

The starting value of the variable may be given below $\sum$ as well. In this case, the summation notation below tells us that the initial value of $x$ will be equal to 1.  

```{marginfigure}
$$\sum_{x=1}$$
```
  
In some cases, we may also wish to designate an ending value of the variable, which we can include above the $\sum$.  The summation notation below also tells us that the final value of $x$ will be equal to 5.  

```{marginfigure}
$$\sum_{x=1}^5$$
```

In all dealings with summation notation, variables will only take integer values, beginning and ending at any values provided within the summation notation.  Note that some summations may use an "ending" value of $\infty$, which would involve the summation of an infinite number of values.  
 
Let's look at a basic summation problem.  

```{marginfigure}
$$\sum_{x=1}^5 2x$$
```

The summation above means that we will take the $x$ values, starting at $x=1$, and multiply the value of $x$ by 2.  We will continue to do this for each integer value of $x$ until we have reached our ending value of 5.  Then we will sum all of our results (five of them, in this case) to produce one final value for the summation.  

```{marginfigure}
$$
\begin{split}
\begin{aligned}
\sum_{x=1}^5 2x\\
& =2*1+2*2+2*3+2*4+2*5\\
& =2+4+6+8+10\\
& =30
\end{aligned}
\end{split}
$$
```

Summation can be calculated over a variety of algebraic expressions.  Another, perhaps more challenging, example is shown below.  

$$
\begin{split}
\begin{aligned}
\sum_{x=0}^3 x^2-4x+1\\
& =(0^2-4*0+1)+(1^2-4*1+1)+(2^2-4*2+1)+(3^2-4*3+1)\\
& =1+(-2)+(-3)+(-2)\\
& =-6
\end{aligned}
\end{split}
$$

## Notation in LaTeX

LaTex is a common typesetting used to express mathematical notation.  The summation notation symbols thus far in this text have been written using LaTex.  Inline LaTeX commands can be added by including a \$ symbol on either side of the LaTeX command.  To create entire LaTeX code chunks, include two \$$ symbols before and after the chunk of LaTeX.

## Sums

To display the $\sum$ symbol in LaTex, use the command `\sum_{lower}^{upper}`.  

The sum expression can be added using the command:  
```{marginfigure}
$$\sum_{lower}^{upper}$$ 
```

Sum limits can be written to appear above and below the operator.  

```{marginfigure}
$$
\sum\limits_{t=0}^{n}\dfrac{CF_t}{(1+r)^t}
$$
```

## Products

The multiplication or product over a sequence is denoted as the uppercase Greek letter pi, $\prod$.  The product command can be added using the expression `\prod_{x = a}^{b} f(x)`

```{marginfigure}
$$\prod_{x = a}^{b} f(x)$$
```

The lower and upper limits of summation after the `\sum` are both optional.  Products and integrals work similarly, only with `\prod` and `\int`:

Delimiters, like parentheses or braces, can automatically re-size to match what they are surrounding. This is done by using `\left` and `\right`:

$$\left( \sum_{i=1}^{n}{i} \right)^2 = \left( \frac{n(n-1)}{2}\right)^2 = \frac{n^2(n-1)^2}{4}$$

## Summary of Mathematical Notations

Below are some common mathematical functions that are often used. 

1. x = y
```{marginfigure}
$$1. x = y$$
```

2. x < y

$$2. x < y$$
3. x > y

$$3.x > y$$
4. x `\le` y
$$4.x \le y$$
5. x `\ge` y 
$$5.x \ge y$$
6. x`^{n}`
$$6.x^{n}$$
7. `x_{n}`
$$7.x_{n}$$
8. `overline{x}`
$$8.\overline{x}$$
9. `hat{x}`
$$9.\hat{x}$$
10. `tilde{x}`
$$10.\tilde{x}$$
11. `frac{a}{b}`
$$11.\frac{a}{b}$$
12. `binom{n}{k}`
$$12.\binom{n}{k}$$
13. `x_{1} + x_{2} + \cdots + x_{n}`
$$13.x_{1} + x_{2} + \cdots + x_{n}$$
14. `|A|`
$$14.|A|$$
15. `x \in A`
$$15.x \in A$$
16. `x \subset B`
$$16.x \subset B$$
17. `x \subseteq B`
$$17.x \subseteq B$$
18. `A \cup B`
$$18.A \cup B$$
19. `{1, 2, 3\}`
$$19.\{1, 2, 3\}$$
20. `\sin(x)`
$$20.\sin(x)$$
21. `\log(x)`
$$21.\log(x)$$
22. `\int_{a}^{b}`
$$22.\int_{a}^{b}$$
23. `\left(\int_{a}^{b} f(x) \; dx\right)`
$$23.\left(\int_{a}^{b} f(x) \; dx\right)$$
24. `\left. F(x) \right|_{a}^{b}`
$$24.\left. F(x) \right|_{a}^{b}$$
25. `\sum_{x = a}^{b} f(x)`
$$25.\sum_{x = a}^{b} f(x)$$
26. `\prod_{x = a}^{b} f(x)`
$$26.\prod_{x = a}^{b} f(x)$$

## Sequences and Summation Notation

Often, especially in the context of optimization, summation notation is used to find the sum of a sequence of terms.  The summation below represents a summing of the first five values of of a sequence of the variable $x$.  The location of the values within the sequence are given by an index value, $i$ in this case.
$$\sum_{i=1}^5 x_i=x_1+x_2+x_3+x_4+x_5$$

Coefficient values may also be included in a summation, as shown below.
```{marginfigure}
$$\sum_{i=1}^5 (10-i)x_i=9x_1+8x_2+7x_3+6x_4+5x_5$$
```

## Applications of Summation

Sequence applications of summation notation can be very practical in that we can extract real-world data values given in an array or a matrix.

For example, let's imagine that the itemized cost for the production of a product from start to finish is that which is given in the table below.

```{r}

##Load Library
library(pander,quietly = TRUE)

#Load Data Table
M1<-matrix(c(11,12,13,14,15), ncol=1)
rownames(M1)<-list("Design", "Materials", 
                   "Production", "Packaging", "Distribution")
colnames(M1)<-list("Product 1")
pander(M1, 
       caption="Table 1: Itemized Production Costs for Product 1")

```

We might wish to determine the total cost to produce the product from start to finish.  We can extract the data from our cost matrix and use summation to find the total cost.

$$
\begin{split}
\begin{aligned}
&\sum_{i=1}^5x_i\\
&=x_1+x_2+x_3+x_4+x_5\\
&=11+12+13+14+15\\
&=65
\end{aligned}
\end{split}
$$

Let's say we now have additional products being produced.

```{r}

#Load Data Table 2
M2<-matrix(c(11,12,13,14,15,21,22,23,24,25,31,32,33,34,35), ncol=3)
rownames(M2)<-list("Design", "Materials", "Production", "Packaging", "Distribution")
colnames(M2)<-list("Product 1", "Product 2", "Product 3")
pander(M2, caption="Table 2: Itemized Production Costs for Three Products")

```

We might wish to determine the cost to produce each of the three products from start to finish.  We could show this with the following summation notation.

$$\sum_{i=1}^5x_{i,j}\ \forall\ j
$$

This notation indicates that we are summing the cost values in the $i$ rows for each product in column $j$.  Note that the symbol $\forall$ shown in the summation above translates to the phrase"for all."  The summation expression above can be interpreted as "the sum of all values of $x_{i,j}$, starting with an initial value of $i=1$, $for\ all$ values of $j$."  The expression will result in $j$ summations.

$$
\begin{split}
\begin{aligned}
&\sum_{i=1}^5x_{i,j}\ \forall \ j\\
&=x_{1,1}+x_{2,1}+x_{3,1}+x_{4,1}+x_{5,1}\\
&=11+12+13+14+15\\
&=65 &\text{Cost for Product 1}\\
\\
AND\\
\\
&=x_{1,2}+x_{2,2}+x_{3,2}+x_{4,2}+x_{5,2}\\
&=21+22+23+24+25\\
&=115 &\text{Cost for Product 2}\\
\\
AND\\
\\
&=x_{1,3}+x_{2,3}+x_{3,3}+x_{4,3}+x_{5,3}\\
&=31+32+33+34+35\\
&=165 &\text{Cost for Product 3}
\end{aligned}
\end{split}
$$

We can see that the summation expression resulted in three summation values since $j$, the number of products, is equal to three.  These summation values are 65, 115, and 165, representing the total cost from start to finish to produce Product 1, Product 2, and Product 3 respectively.

## Double Summation

For some projects or models, we may need to add one summation into another. This procedure is called "double summation."  Consider the following double summation expression:  `\sum_{i=1}^3\sum_{j=1}^4 (i+j)`

```{marginfigure}
$$\sum_{i=1}^3\sum_{j=1}^4 (i+j)$$
```

Note that the expression contains two $\sum$ symbols, indicating a double summation.  The double summation would expand as shown below.

```{marginfigure}
$$
\begin{split}
\begin{aligned}
\sum_{i=1}^3\sum_{j=1}^4 (i+j)\\
&=(1+1)+(2+1)+(3+1)\\
&+(1+2)+(2+2)+(3+2)\\
&+(1+3)+(2+3)+(3+3)\\
&+(1+4)+(2+4)+(3+4)
\end{aligned}
\end{split}
$$
```

## Applications of Double Summation

Consider a transportation application using double summation in which we want to ship a given amount of product $X$ from location $i$ to location $j$, denoted $X_{i,j}$.  The summation notation for this application is shown below.

```{marginfigure}
$$\sum_{i=1}^n\sum_{j=1}^m X_{i,j}$$
```

Expanding on the previous summation, we may also want to include shipping costs $C$ from location $i$ to location $j$, denoted $C_{i,j}$.  Combining the amount of product with the related shipping costs would result in the double summation expression shown below.

$$\sum_{i=1}^n\sum_{j=1}^m C_{i,j}X_{i,j}$$

## Exercises

1.  Write all terms and calculate the summation for each exercise.

$$
\begin{split}
\begin{aligned}
A. && \sum_{x=1}^4 x+3\\
\\
B.  && \sum_{x=0}^5 8x-1\\
\end{aligned}
\end{split}
$$

2.  Write as a sum of all terms in the sequence.

$$
\begin{split}
\begin{aligned}
\sum_{x=1}^6 (2i)x
\end{aligned}
\end{split}
$$

3.  Write a summation to represent the total cost associated with producing three items of Product 1.  Use the values from Table 1 to evaluate your summation expression.

4.  Write all terms and calculate the summation for each exercise.

$$
\begin{split}
\begin{aligned}
A. && \sum_{i=1}^3\sum_{j=1}^4 (i*j)\\
\\
B.  && \sum_{i=1}^5\sum_{j=1}^2 (3i-j)\\
\end{aligned}
\end{split}
$$

5.  A company compensates its employees with an annual salary and an annual bonus.  

A.  Write an expression using summation notation to represent the total annual compensation $i$ (including salary and bonus) for each job title $j$.  

B.  Write a double summation expression to represent the total amount the company pays annually to compensate all its employees if each job title has $n_j$ employees.  
(Need possible sample expression for A. and B.)

## References
http://www.columbia.edu/itc/sipa/math/summation.html
https://www.overleaf.com/learn/latex/Integrals,_sums_and_limits#Sums_and_products 
http://www.u.arizona.edu/~kuchi/Courses/MAT167/Files/LH_LEC.0130.Intro.SummationNotation.pdf
https://www.bates.edu/mathematics/resources/what-is-latex/
http://math.hws.edu/gassert/LaTeX_Guide_Title.pdf 
http://www.statpower.net/Content/310/R%20Stuff/SampleMarkdown.html
http://www.calvin.edu/~rpruim/courses/m343/F12/RStudio/LatexExamples.html
http://pages.stat.wisc.edu/~jgillett/371/RStudio/RMarkdown.pdf
https://www.academia.edu/21823064/Creating_Dynamic_Mathematical_Documents_in_RStudio_by_Unifying_Computing_with_R_and_Typesetting_with_LaTeX
