---
title: "Appendix: Table Formats"
author: "Timothy R. Anderson"
header-includes:
- \usepackage{longtable}
- \usepackage{caption}
monofont: Times New Roman
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
    toc: TRUE
    number_sections: true
    highlight: monochrome
  tufte::tufte_html: default
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
---

\pagestyle{headings}

```{r Ch95setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy = F)
library (tufte)
library (pander)
library (tidyr)
library(magrittr)
library(poliscidata)
library(coronavirus)
library(tidyverse)
library(devtools)
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library (tint, quietly = TRUE))
    # Loading this last masks margin_note, newthought,
    #    quote_footer, sans_serif from the tufte package.

# Packages for Tables
library (pander)
library (xtable)
library (huxtable)
library (kableExtra)

```

# Table Formats

## Overview

Displaying of tables is critical in quantitative work such as optimization and more generally analytics.  The goal of this Appendix is to demonstrate some of the common packages for displaying tables, in one place as a part of the book within the context of its own formatting.  

For data, let's download data on current Covid-19 infections.

```{r}
library(coronavirus)
data(coronavirus)
require(dplyr)
# Get top confirmed cases by state
coronavirus %>%
  filter(type == "confirmed") %>%
  group_by(country) %>%
  summarise(total = sum(cases)) %>%
  arrange(-total) %>%
  head(20)

refresh_coronavirus_jhu()

coronavirus %>%
  filter(type == "confirmed") %>%
  group_by(country) %>%
  summarise(total = sum(cases)) %>%
  arrange(-total) %>%
  head(20)

```


```{r}
# Get the number of recovered cases in China by province
coronavirus %>%
filter(type == "recovered", country == "US") %>%
group_by(province) %>%
summarise(total = sum(cases)) %>%
arrange(-total)

```

```{r}
library (tidycovid19)
library (tidyverse)

df <- download_jhu_csse_covid19_data(
  type = "us_county", silent = TRUE, cached = TRUE
)
df1<-df %>%
  dplyr::filter(!is.na(state)) %>%
  dplyr::group_by(state) %>%
  dplyr::summarise( deaths = max(deaths, na.rm = TRUE)) %>%
  dplyr::arrange(-deaths) %>%
  dplyr::top_n(10)

# Calculate Cumulative totals by state
df2<-df %>%            # Calculate Cumulative totals by state
  dplyr::filter(!is.na(state)) %>%
  dplyr::group_by(county, state) %>%
  dplyr::summarise(deaths = max(deaths, na.rm = TRUE), 
                   confirmed = max(confirmed, na.rm = TRUE)) %>%
  dplyr::arrange(-deaths) 

df3<-df2 %>%
  dplyr::group_by(state) %>%
  dplyr::summarise(deaths = sum(deaths, na.rm = TRUE), 
                   confirmed = sum(confirmed, na.rm = TRUE)) %>%
  dplyr::arrange(-deaths)


df4<-df3 %>%   # Convert State name to format used by poliscidata package
  mutate(state=str_replace(state, "New Hampshire", "NewHampshire")) %>%
  mutate(state=str_replace(state, "New Jersey", "NewJersey")) %>%
  mutate(state=str_replace(state, "New Mexico", "NewMexico")) %>%
  mutate(state=str_replace(state, "New York", "NewYork")) %>%
  mutate(state=str_replace(state, "North Carolina", "NorthCarolina")) %>%
  mutate(state=str_replace(state, "South Carolina", "SouthCarolina")) %>%
  mutate(state=str_replace(state, "North Dakota", "NorthDakota")) %>%
  mutate(state=str_replace(state, "South Dakota", "SouthDakota")) %>%
  mutate(state=str_replace(state, "West Virgina", "WestVirgina"))
  
head(df4)

# Remove non-states from dataset
df5 <- df4[!(df4$state %in% c("Puerto Rico", "District of Columbia", "Guam","Virgin Islands", "Grand Princess", "Northern Mariana Islands", "American Samoa", "Diamond Princess" )),]

# head(df5)
# tbl_states <- as.tibble(states)
# df6 <- left_join (df5, tbl_states,by=c("state"))
# head(df6[,1:6])
```


```{r pander_table}
pander (head(df5), caption = "Pander generated table.")
```

```{r kable_table}
kable(df5, caption="Kable generated table.")
```


```{r huxtable_table}
huxtable(df5, caption="Huxtable generated table.")
```
