---
title: "ETM 540 - Practice Problems"
author: "Viktor E. Viking"
date: "`r Sys.Date()`"
output: 
  pdf_book:
    number_sections: true
---

```{r, include=FALSE, eval=FALSE}
# This RMD makes use of several items:
# The above header includes a dynamic date based on when the file is rendered.
# The output format turns off the automatic number of sections so that
#     sections have names inline with parts of the homework.
# This code chunk uses options for important purposes:
#    include=FALSE to not show the code or messages
#    eval=FALSE so that is is only run when the code chunk is run directly -
#         not when all codes are run or the file is knitted.
#         This option could be used when a code chunk error is preventing a
#         a file from being rendered to PDF.
#         Another option is to set error=TRUE to allow the file to be 
#         knitted even when there is an error in that code chunk.
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy = FALSE)
knitr::opts_chunk$set(cache = FALSE)
library(bookdown); library(rmarkdown); 
rmarkdown::render("_Practice Problems.Rmd", "pdf_book")
```

The following two formulation problems are given without detailed data.  This is done to help keep people from getting fixated or distracted by the data.  Once a formulation is made, it can be implemented with _any_ specific data.


# Ship Loading

During Covid, international shipping has had a a lot of disruption. You have five food commodities available to load on a ship. Each food commodity, _i_, has the following characteristics:

- $A_{i}$, Amount of commodity _i_ available in tons, 
- $V_{i}$, Volume of commodity _i_ in cubic meters per ton, and
- $P_{i}$, Profit per ton for commodity _i_. 

The ship has four holds: Front, Left, Right, Back.  Let's denote hold by _h_ and each hold has capacity limits by weight and volume.

- $W_{h}$, Capacity of hold _h_ by weight (tons) and
- $U_{h}$, Capacity of hold _h_ by volume in cubic meters

For safety  on the seas, the ship must be well balanced.

- The weight of the front hold's load cannot be more than 110% of the back hold's load.
- The weight of the back hold's load cannot be more than 120% of the front hold's load.
- The left and right hold must be within 15% of the other's weight.
- The two middle holds combined must be between 30% and 60% of the total weight of all loads.

## Formulate an optimization model

> Assume food commodities can be mixed in a hold.

### Define appropriate decision variables

$x_{i,h}$ is the amount of food commodity i in hold h.

### Create an objective function

maximize profit =  $\text{Max:} \sum_{h=1}^{4} \sum_{i=1}^{5} P_{i,j} x_{i,j} \\$



### Create appropriate constraints

Use no more of a commodity i than we have available.
$\sum_{h=1}^{4} x_{i,h} \leq A_i  \forall \; i$

Hold weight capacity
$\sum_{i=1}^{5} x_{i,h} \leq W_h  \forall \; h$

Hold volume capacity
$\sum_{i=1}^{5} V_i \cdot x_{i,h} \leq U_h  \forall \; h$

The weight of the front hold's load cannot be more than 110% of the back hold's load.
$\sum_{i=1}^{5} x_{i,FRONT} \leq 1.1 \cdot \sum_{i=1}^{5} x_{i,BACK}$

or better yet define front is h=1, right is h=2, back is h=3, left is h=4

$\sum_{i=1}^{5} x_{i,1} \leq 1.1 \cdot \sum_{i=1}^{5} x_{i,3}$


The weight of the back hold's load cannot be more than 120% of the front hold's load.
$\sum_{i=1}^{5} x_{i,3} \leq 1.2 \cdot \sum_{i=1}^{5} x_{i,1}$



The left and right hold must be within 15% of the other's weight.
$\sum_{i=1}^{5} x_{i,2} \leq 1.15 \cdot \sum_{i=1}^{5} x_{i,4}$
$\sum_{i=1}^{5} x_{i,4} \leq 1.15 \cdot \sum_{i=1}^{5} x_{i,2}$


The two middle holds combined must be between 30% and 60% of the total weight of all loads.
$\sum_{i=1}^{5} x_{i,2} + \cdot \sum_{i=1}^{5} x_{i,4}\geq 0.3 \cdot \sum_{h=1}^{4} \sum_{i=1}^{5}  x_{i,h}$
$\sum_{i=1}^{5} x_{i,2} + x_{i,4}\geq 0.3 \cdot \sum_{h=1}^{4} \sum_{i=1}^{5}  x_{i,h}$

$\sum_{i=1}^{5} x_{i,2} + x_{i,4}\leq 0.6 \cdot \sum_{h=1}^{4} \sum_{i=1}^{5}  x_{i,h}$



## Formulate an optimization model where crops cannot be mixed in a hold.

> Hint:  We wouldn't want to mix rice and wheat as separating the two could be quite time consuming.  Therefore, only one crop can be in a hold.  

### Define appropriate decision variables

Add to the previous model

$y_{i,h} = 1$ if crop i is in hold h. 0 otherwise.

### Create an objective function

### Create appropriate constraints

summation over i of y_ih =1 for all h

x_ih <= M * y_ih for all ih
a good big M value is A_i or W_h


# Placement of Emergency Response Teams

The City of Bartlandia wants to locate two emergency response teams to serve residents in need.  The city can be thought of as a grid of six regions with a corresponding time to travel from neighborhood _i_ to neighborhood _j_ of $T_{i,j}$.  The neighborhoods have populations of $P_j$. They would like to place the two teams so as maximize the number of people that can a team can respond to within five minutes.

## Formulate an appropriate optimization model

### Define the Decision Variables

Let's define our variables.

- $x_i$ is the decision whether to place a team in region i.  $x_i=1$ if a team is in location i.
- $y_j$ is the decision whether region j is served by one or more teams.  Therefore, $y_j=1$ if a region j is served.



### Define Any Additional Datasets Needed

Let's define $R_{i,j}=1 if a team in region i can reach region j in the 5 minute requirement.

### Create the Optimizaton Model

$$
 \begin{split}
 \begin{aligned}
    \text{Max:}  &  \sum_{j=1}^{6} P_j y_j \\
    \text{subject to } & y_j \leq \sum_{i=1}^{6} R_{i,j}x_i  \; \forall \; j\\
                       & x_i,y_j   \in \{0,1\}  \; \forall \; i, j
  \end{aligned}
  \end{split}
$$
## For the following data, create and implement a model.  

### Load Appropriate Packages

```{r, warning=FALSE, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy = FALSE)
knitr::opts_chunk$set(cache = FALSE)
library (dplyr, quietly = TRUE)
library (ROI, quietly = TRUE)
library (ROI.plugin.glpk, quietly = TRUE)
library (ompr, quietly = TRUE)
library (ompr.roi, quietly = TRUE)
library (knitr)
library (kableExtra)
```

### Prepare the Data



### Build the Model


### Solve the Model


## Present and interpret the results

```{r, error=TRUE}
dsfdsfsdfsddf
```
I've tried fixing it but I can't find the problem.

